---
title: "problem3"
author: "Varun Venkatesh"
date: "`r Sys.Date()`"
output: html_document
---
Problem 3: Price Elasticity of Demand
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Read in the data
milk_data <- read.csv("milk.csv")

# Examine the data
head(milk_data)
summary(milk_data)

# Plot the relationship between price and quantity
plot(milk_data$price, milk_data$sales, 
     main="Relationship between Milk Price and Sales",
     xlab="Price ($)", ylab="Sales (number of participants)", 
     pch=19, col="darkblue")
```
The power-law model for price elasticity is given by:
$Q = KP^{\beta}$
Where:

$Q$ is the quantity demanded (sales)
$P$ is the price
$K$ is a constant
$\beta$ is the price elasticity of demand

Taking the natural logarithm of both sides:
$\log(Q) = \log(K) + \beta \log(P)$
This transforms the model into a linear relationship in the log-log space, where $\beta$ is the slope and represents the price elasticity.


```{r cars}
# Transform to log-log scale
milk_data$log_price <- log(milk_data$price)
milk_data$log_sales <- log(milk_data$sales)

# Fit the linear model in log-log space
elasticity_model <- lm(log_sales ~ log_price, data = milk_data)
summary(elasticity_model)

# Extract the price elasticity (the coefficient of log_price)
price_elasticity <- coef(elasticity_model)[2]
cat("Estimated price elasticity of demand for milk:", round(price_elasticity, 3), "\n")

# Plot the relationship in log-log scale
plot(milk_data$log_price, milk_data$log_sales, 
     main="Log-Log Plot of Milk Price and Sales",
     xlab="Log Price", ylab="Log Sales", 
     pch=19, col="darkblue")
abline(elasticity_model, col="red", lwd=2)
```

Now let's compute a 95% bootstrap confidence interval for the price elasticity:


```{r pressure, echo=FALSE}
library(boot)

# Bootstrap function for elasticity
boot_elasticity <- function(data, indices) {
  boot_sample <- data[indices, ]
  model <- lm(log_sales ~ log_price, data = boot_sample)
  return(coef(model)[2])  # return the slope (elasticity)
}

# Set random seed for reproducibility
set.seed(123)

# Perform bootstrap
boot_results <- boot(milk_data, boot_elasticity, R = 1000)

# Calculate 95% confidence interval
ci_elasticity <- boot.ci(boot_results, type = "perc")

# Display results
cat("Price elasticity of demand for milk:", round(price_elasticity, 3), "\n")
cat("95% Bootstrap CI: [", round(ci_elasticity$percent[4], 3), ",", 
    round(ci_elasticity$percent[5], 3), "]\n")


```

Results and Interpretation

To estimate the price elasticity of demand for milk, I used the power-law model $Q = KP^{\beta}$, where $\beta$ represents the price elasticity. By taking the natural logarithm of both sides, I transformed this into a linear model in log-log space: $\log(Q) = \log(K) + \beta \log(P)$. Fitting this linear model to the data, I found the estimated price elasticity of demand for milk to be -1.619  with a 95% bootstrap confidence interval of [ -1.771 , -1.439 ]
Since this elasticity is negative, it confirms the law of demand: as price increases, quantity demanded decreases. The magnitude of the elasticity is greater than 1, which indicates that milk demand is elastic,consumers are relatively sensitive to price changes. Specifically, a 1% increase in the price of milk leads to approximately a 1.619 % decrease in quantity demanded.
