---
title: "Problem 1"
author: "Varun Venkatesh"
date: "`r Sys.Date()`"
output: html_document
---

```{r problem 1, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Read in the data
creatinine_data <- read.csv("creatinine.csv")

# Look at the first few rows
head(creatinine_data)

# Basic summary statistics
summary(creatinine_data)

# Fit linear regression model
creat_model <- lm(creatclear ~ age, data = creatinine_data)

# View the model summary
summary(creat_model)

#fit a linear regression model to the data:

# Plot the data and the regression line
plot(creatinine_data$age, creatinine_data$creatclear, 
     xlab = "Age (years)", ylab = "Creatinine Clearance Rate (mL/minute)",
     main = "Creatinine Clearance Rate vs. Age",
     pch = 19, col = "darkblue")
abline(creat_model, col = "red", lwd = 2)
```

## Problem 1

A)  What creatinine clearance rate should we expect for a 55-year-old?

```{r cars}
# Extract coefficients from the model
intercept <- coef(creat_model)[1]
slope <- coef(creat_model)[2]

# Calculate expected creatinine clearance for a 55-year-old
expected_55 <- intercept + slope * 55
expected_55

  
```
The expected creatinine clearance rate for a 55-year-old is 
113.7 ml/minute

Explanation: I calculated this using the linear regression equation: $\hat{y} = \beta_0 + \beta_1 \times age$, where $\hat{y}$ is the predicted creatinine clearance rate, $\beta_0$ is the intercept, and $\beta_1$ is the slope coefficient for age. Substituting 55 for age gives the expected clearance rate.

## Including Plots

B)  How does creatinine clearance rate change with age?

```{r pressure, echo=FALSE}
# The slope coefficient represents how creatinine clearance changes with age
slope

```

The creatinine clearance rate changes by -0.62 mL/minute per year of age. Explanation: The slope coefficient from our linear regression model directly tells us how much creatinine clearance rate changes for each one-year increase in age. The negative value indicates that creatinine clearance rate decreases with age.

C)  Comparing creatinine clearance rates adjusted for age We need to determine which patient has a higher clearance rate relative to what would be expected for their age.

```{r}
# Calculate expected clearance rate for a 40-year-old
expected_40 <- intercept + slope * 40
# Calculate expected clearance rate for a 60-year-old
expected_60 <- intercept + slope * 60

# Calculate residuals (observed - expected)
residual_40 <- 135 - expected_40
residual_60 <- 112 - expected_60

# Show results
expected_40
expected_60
residual_40
residual_60



```

##Comparison:

40-year-old: Observed rate = 135 mL/minute, Expected rate = 123.02  mL/minute, Difference = 11.98 mL/minute 60-year-old: Observed rate = 112 mL/minute, Expected rate =   110.6 mL/minute, Difference = 1.38 mL/minute

The r ifelse(residual_40 \> residual_60, "40-year-old", "60-year-old") has a healthier (higher) creatinine clearance rate for their age. Explanation: To determine which patient has a healthier rate for their age, I compared each patient's observed rate to their age-expected rate (calculated from our regression model). The patient with a larger positive difference (residual) has a clearance rate that exceeds expectations for their age by a greater amount, indicating a relatively healthier kidney function.
