---
title: "problem2file"
author: "Varun Venkatesh"
date: "`r Sys.Date()`"
output: html_document
---
Load Data
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Read in the data
covid_data <- read.csv("covid.csv")

# Examine the data
head(covid_data)
str(covid_data)
summary(covid_data)

# Check data for both countries
table(covid_data$country)

```

Fitting Exponential Growth Models
For exponential growth models, use the formula:

$\text{deaths} = \text{deaths}_0 \times e^{rt}$

where $r$ is the growth rate and $t$ is time (days since first death).

Taking the log of both sides:

$\log(\text{deaths}) = \log(\text{deaths}_0) + rt$

Can fit a linear model to the log of deaths:


```{r cars}
# Create log of deaths
covid_data$log_deaths <- log(covid_data$deaths)

# Splitting the data by country
italy_data <- subset(covid_data, country == "Italy")
spain_data <- subset(covid_data, country == "Spain")

# Fit exponential models (linear in log-scale)
italy_model <- lm(log_deaths ~ days_since_first_death, data = italy_data)
spain_model <- lm(log_deaths ~ days_since_first_death, data = spain_data)

# Model summaries
summary(italy_model)
summary(spain_model)

# Extract growth rates
growth_rate_italy <- coef(italy_model)[2]
growth_rate_spain <- coef(spain_model)[2]

# Calculate doubling times (ln(2)/r)
doubling_time_italy <- log(2)/growth_rate_italy
doubling_time_spain <- log(2)/growth_rate_spain

# Display results
cat("Growth rate (Italy):", round(growth_rate_italy, 3), "\n")
cat("Doubling time (Italy):", round(doubling_time_italy, 1), "days\n")
cat("Growth rate (Spain):", round(growth_rate_spain, 3), "\n")
cat("Doubling time (Spain):", round(doubling_time_spain, 1), "days\n")
```


Bootstrap Confidence Intervals
Compute 95% bootstrap confidence intervals for the growth rates and doubling times:

```{r pressure, echo=FALSE}
library(boot)

# Bootstrap function for growth rate
boot_growth_rate <- function(data, indices) {
  boot_sample <- data[indices, ]
  model <- lm(log_deaths ~ days_since_first_death, data = boot_sample)
  return(coef(model)[2])  # return the slope (growth rate)
}

# Bootstrap function for doubling time
boot_doubling_time <- function(data, indices) {
  boot_sample <- data[indices, ]
  model <- lm(log_deaths ~ days_since_first_death, data = boot_sample)
  return(log(2)/coef(model)[2])  # return doubling time
}

# Set random seed for reproducibility
set.seed(123)

# Perform bootstrap for Italy
boot_italy_growth <- boot(italy_data, boot_growth_rate, R = 1000)
boot_italy_doubling <- boot(italy_data, boot_doubling_time, R = 1000)

# Perform bootstrap for Spain
boot_spain_growth <- boot(spain_data, boot_growth_rate, R = 1000)
boot_spain_doubling <- boot(spain_data, boot_doubling_time, R = 1000)

# Calculate 95% confidence intervals
ci_italy_growth <- boot.ci(boot_italy_growth, type = "perc")
ci_italy_doubling <- boot.ci(boot_italy_doubling, type = "perc")
ci_spain_growth <- boot.ci(boot_spain_growth, type = "perc")
ci_spain_doubling <- boot.ci(boot_spain_doubling, type = "perc")

# Display results
cat("Italy growth rate:", round(growth_rate_italy, 3), 
    "95% CI: [", round(ci_italy_growth$percent[4], 3), ",", 
    round(ci_italy_growth$percent[5], 3), "]\n")

cat("Italy doubling time:", round(doubling_time_italy, 1), "days", 
    "95% CI: [", round(ci_italy_doubling$percent[4], 1), ",", 
    round(ci_italy_doubling$percent[5], 1), "]\n")

cat("Spain growth rate:", round(growth_rate_spain, 3), 
    "95% CI: [", round(ci_spain_growth$percent[4], 3), ",", 
    round(ci_spain_growth$percent[5], 3), "]\n")

cat("Spain doubling time:", round(doubling_time_spain, 1), "days", 
    "95% CI: [", round(ci_spain_doubling$percent[4], 1), ",", 
    round(ci_spain_doubling$percent[5], 1), "]\n")
```


Line graph showing reported daily deaths over time for both countries:
```{r}
library(ggplot2)

# Create a line graph
ggplot(covid_data, aes(x = days_since_first_death, y = deaths, color = country)) +
  geom_line(size = 1) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("Italy" = "darkblue", "Spain" = "darkred")) +
  labs(title = "COVID-19 Daily Deaths in Italy and Spain",
       x = "Days Since First Death",
       y = "Number of Deaths",
       color = "Country") +
  theme_minimal() +
  theme(legend.position = "top")
```

Printed Summary of results:
```{r}
# Create a summary data frame using existing variable values
results_summary <- data.frame(
  Country = c("Italy", "Spain"),
  Growth_Rate = c(growth_rate_italy, growth_rate_spain),
  Growth_Rate_CI_Lower = c(ci_italy_growth$percent[4], ci_spain_growth$percent[4]),
  Growth_Rate_CI_Upper = c(ci_italy_growth$percent[5], ci_spain_growth$percent[5]),
  Doubling_Time_Days = c(doubling_time_italy, doubling_time_spain),
  Doubling_Time_CI_Lower = c(ci_italy_doubling$percent[4], ci_spain_doubling$percent[4]),
  Doubling_Time_CI_Upper = c(ci_italy_doubling$percent[5], ci_spain_doubling$percent[5])
)

# Print formatted results
print(results_summary)

# For nicer formatting in RMarkdown
library(knitr)
kable(results_summary, 
      col.names = c("Country", "Growth Rate", "GR CI Lower", "GR CI Upper", 
                    "Doubling Time (days)", "DT CI Lower", "DT CI Upper"),
      digits = c(0, 3, 3, 3, 1, 1, 1))

# Optional: Create a formatted string for each country that could be used in text
italy_text <- sprintf("Italy: Growth rate: %.3f (95%% CI: [%.3f, %.3f]), Doubling time: %.1f days (95%% CI: [%.1f, %.1f])",
                     growth_rate_italy, 
                     ci_italy_growth$percent[4], ci_italy_growth$percent[5],
                     doubling_time_italy,
                     ci_italy_doubling$percent[4], ci_italy_doubling$percent[5])

spain_text <- sprintf("Spain: Growth rate: %.3f (95%% CI: [%.3f, %.3f]), Doubling time: %.1f days (95%% CI: [%.1f, %.1f])",
                     growth_rate_spain, 
                     ci_spain_growth$percent[4], ci_spain_growth$percent[5],
                     doubling_time_spain,
                     ci_spain_doubling$percent[4], ci_spain_doubling$percent[5])
```
The growth rate represents the rate at which the number of daily deaths was increasing exponentially during this period. A higher growth rate means faster exponential growth. The doubling time is the time it takes for the number of daily deaths to double, calculated as ln(2)/growth rate.
From our analysis, Spain appears to have had a slightly higher growth rate than Italy during this initial wave, resulting in a shorter doubling time for daily deaths. This suggests that in the early stages of the pandemic, the daily death count was increasing at a faster pace in Spain compared to Italy.
